Estimating latent means model with regression models in R

### 2023-12-04 manuscript.qmd

---
title: 'Bias no more: Introduction to latent group-mean centering in hierarchical bayesian models'
shorttitle: Latent group-mean centering
leftheader: Latent group-mean centering
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Tilburg University
    email: m.j.vuorre@tilburguniversity.edu
  - name: Author 2
    affiliation: 2
    address: 
    email:

affiliation:
  - id: 1
    institution: Tilburg University
  - id: 2
    institution: Institution 2

authornote: |
  \noindent \textit{This manuscript is not yet peer-reviewed.}

abstract: |
  Researchers studying links between observed variables in longitudinal and other repeated measures data sets have long recognized the importance of person-mean centering in order to isolate within- and between-person associations. However, a critical issue in centering---treating the person means as unknown parameters rather than observed values---has received less attention among applied practitioners. One reason for this has been the general unavailability and perceived complexity of latent person-mean centering in contrast to observed mean centering. Here, we introduce latent person-mean centering and accessible syntax for implementing it in a wide variety of flexible bayesian models with the R package brms. We demonstrate the syntax, and benefits of latent centering, through an example analysis of social media effects on adolescent well-being. In appendices, we further contrast the performance of latent mean centering to observed mean centering via simulations, and show how it extends to more than two levels of analysis. Latent mean centering provides unbiased estimates of within- and between-person associations, and we hope that the availability and flexibility of our proposed approach motivates researchers to apply this method widely in longitudinal and other multilevel modelling scenarios.

keywords: longitudinal data, repeated measures, centering, well-being, technology effects
wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"
bibliography: "references.bib"
floatsintext: yes
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
output: 
  papaja::apa6_pdf:
    number_sections: false
    keep_tex: false
  papaja::apa6_docx: 
    number_sections: false
---

@@

```{r}
#| label: tab-data
d %>% 
  slice(1:3, .by = person) %>% 
  head() %>% 
  apa_table(
    caption = "Six rows of example data", 
    note = "sm = passive social media use; smg = grand-mean centered sm; smb = between-person sm; smw = within-person sm; happyw1 = within-person lagged happiness rating.",
    col_spanners = list(`Original data` = c(1, 5), `Centered values` = c(6, 9)),
    digits = c(0, 0, 0, 0, 0, 0, 1, 1, 1), 
    font_size = "small", 
    span_text_columns = FALSE
  )
```

@@

```{r}
#| eval: false
model <- bf(
  happy ~ b0happy + b1happy*(happy1 - b0happy) + b2happy*(sm - b0sm),
  b0happy ~ 1 + (1 | person),
  b1happy ~ 1 + (1 | person),
  b2happy ~ 1 + (1 | person),
  b0sm ~ 1 + (1 | person),
  nl = TRUE,
  center = FALSE
) +
  gaussian()

# Set priors
p <- prior(normal(30, 15), class = b, coef = Intercept, nlpar = b0sm)

fit <- brm(
  model,
  data = d,
  prior = p,
  cores = MAX_CORES,
  threads = THREADS,
  iter = ITER,
  control = list(adapt_delta = 0.99),
  file = "models/brm-fit-1"
)
summary(fit, prior = TRUE)
```

@@

Conceptualization: MV
Methodology: MV
Software: MV
Validation: MV
Formal Analysis: MV
Data Curation: MV
Writing -- Original Draft: MV
Writing -- Review & Editing: MV
Visualization: MV
Supervision: MV
Project Administration: MV

@@

# Acknowledgements

We thank everyone who contributed to this discussion on the Stan discussion board (https://web.archive.org/web/20230207111825/https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424).

# References

::: {#refs custom-style="Bibliography"}
:::

<!-- Create an appendix with new page, table, and figure numbers. -->

\clearpage
\onecolumn

# Appendix Z: Simulation study

<!-- Prepend table & figure numbers with S and reset counters -->
\setcounter{page}{1}
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

Hereafter be monsters and dragons.

Here, we conduct a small simulation study to verify that the proposed estimation method recovers specified parameter values.

We simulate data from a bivariate model of $x$ and $y$ measured over $i$ timepoints from 1 to $I$ and for $j$ subjects in 1 to $J$. In this simulation, we set $J = 30$ and vary $I$ from 6 to 60. Both variables are gaussian, with $x$ a simple random variable with a person-varying mean $\alpha_x$, and $y$ is affected by its previous value and the previous value of $x$.

\begin{align*}
x_{ij} &\sim \operatorname{Normal}(\alpha_{xj}, \sigma^2_x) \\
y_{ij} &\sim \operatorname{Normal}(
  \alpha_{yj} +
  \beta_{1j}(y_{[i-1]j} - \alpha_{yj}) +
  \beta_{2j}(x_{[i-1]j} - \alpha_{xj}), 
  \sigma^2_y
)
\end{align*}

The second model equation above specifies that the $\beta$s model effects of the within-person deviated variables of lagged y and x, by virtue of subtracting from each their respective person-specific (latent) mean. I hope. Then we specify the rest of the multilevel model as usual, butchering the notation with bars to indicate average or population-level regression coefficients. We also specify matrices of standard deviations S and correlations R to pop into the covariance matrix (as those will be estimated, rather than covariances, by brms/Stan).

\begin{align*}
\begin{bmatrix} 
  \alpha_x \\ \alpha_y \\ \beta_1 \\ \beta_2
\end{bmatrix} &\sim \operatorname{MVN}(
\begin{bmatrix} 
  \bar{\alpha}_x \\ \bar{\alpha}_y \\ \bar{\beta}_1 \\ \bar{\beta}_2
\end{bmatrix}, \mathbf{\Sigma}) \\ \\
\Sigma &= \mathbf{SRS} \\ \\
\mathbf{S} &= \begin{bmatrix} 
  \tau_{\alpha_x} &0 &0 &0 \\ 
  0 &\tau_{\alpha_y} &0 &0\\ 
  0 &0 &\tau_{\beta_1} &0 \\ 
  0 &0 &0 &\tau_{\beta_2}
\end{bmatrix} \\ \\
\mathbf{R} &= 
\begin{bmatrix} 
  1 &\rho_{\alpha_x\alpha_y} &\rho_{\alpha_x\beta_1} &\rho_{\alpha_x\beta_2} \\ 
  \rho_{\alpha_x\alpha_y} &1 &\rho_{\alpha_y\beta_1} &\rho_{\alpha_y\beta_2} \\ 
  \rho_{\alpha_x\beta_1} &\rho_{\alpha_y\beta_1} &1 &\rho_{\beta_1\beta_2} \\ 
  \rho_{\alpha_x\beta_2} &\rho_{\alpha_y\beta_2} &\rho_{\beta_1\beta_2} &1 \\
\end{bmatrix}.
\end{align*}

We then estimate the resulting datasets with two models. The first is an lmer model that substitutes the observed means $\bar{x}_{.j}$ and $\bar{y}_{.j}$ for the latent intercepts $\alpha_x$ and $\alpha_y$, respectively. The second is our main brms model, specified as follows:

```{r}
#| include: true
#| echo: true
model <- bf(
  y ~ alphay + betayy*(ylag - alphay) + betayx*(xlag - alphax),
  alphax + alphay + betayy + betayx ~ 1 + (1 |i| individual),
  nl = TRUE
) +
  gaussian()
```

Note that this model doesn't estimate an outcome model for $x$, which may or may not have consequences. We then went ahead and ran the simulations with 100 repetitions at each row of the simulation design matrix. This only took a few hours on a M1 laptop. We show results in Figure \@ref(fig:fig-sim-res). First, we see that observed mean centering recovers the specified values of $\alpha_x$ well--those are not in the model output but just means of $\bar{x}_{.j}$. The latent means estimates here are more variable across simulation runs, but also recover the mean well when the number of timepoints is larger than about 10. $\alpha_y$ remains biased upward even with a large number of timepoints per person, but not with the latent means model.

The key coefficients $\beta_1$ and $_2$ show that latent mean centering effectively debiases the autocorrelated parameter, even at low numbers of timepoints. Estimates of $\beta_2$ are noisier across simulation runs but both models seem to recover the values well (observed mean model is consistently greater?)

Other notable findings:
- The correlation between intercepts (between-person correlation) $\rho$ is massively downward biased?
- The between-person standard deviations are all over the place.

```{r}
# Run simulations
# Save simulation results to file for every nr_individuals * nr_time
dir.create("sims", FALSE)

# Load function to simulate data
source("Generation_Code.r")

# Specify population level parameters
pop <- tibble(
  sigma_x = 1.0,
  sigma_y = 1.0,
  cor_xy = 0.0,
  alpha_x = 0.0,
  alpha_y = 0.0,
  tau_alpha_x = 0.5,
  tau_alpha_y = 0.5,
  cor_alpha = 0.2,
  beta_yy1 = 0.3,
  beta_yx = 0.25,
  beta_xy = 0.0,
  beta_xx1 = 0.0,
  tau_beta_yy1 = 0.1,
  tau_beta_yx = 0.2,
  tau_beta_xy = 0.0,
  tau_beta_xx1 = 0.0,
  cor_beta = 0.0
)

# Save empty brms model
p <- prior(normal(0, 1), class = b, coef = Intercept, nlpar = alphax) +
  prior(normal(0, 1), class = b, coef = Intercept, nlpar = alphay) +
  prior(normal(0, 1), class = b, coef = Intercept, nlpar = betayy) +
  prior(normal(0, 1), class = b, coef = Intercept, nlpar = betayx) +
  prior(lkj(2), class = cor, group = individual) +
  prior(student_t(7, 0, 1), class = sd, group = individual, nlpar = alphax) +
  prior(student_t(7, 0, 1), class = sd, group = individual, nlpar = alphay) +
  prior(student_t(7, 0, 0.5), class = sd, group = individual, nlpar = betayy) +
  prior(student_t(7, 0, 0.5), class = sd, group = individual, nlpar = betayx)
fit_brms_empty <- brm(
  model,
  data = simulate_twolevel_data()$data[[1]],
  prior = p,
  control = list(adapt_delta = 0.95),
  chains = 0,
  file = "models/brm-fit-empty"
)

# Function to estimate parameters
simulate_estimate <- function(nr_indiv = 30, nr_time = 10, iter = 2) {
  path <- str_glue("sims/nr_indiv-{nr_indiv}__nr_time-{nr_time}.rds")
  if (!file.exists(path)) {

    library(furrr)
    plan(multisession, workers = 8)

    sims <- pop %>% 
      # Cross design with sample sizes and simulation iterations
      crossing(
        nr_indiv = nr_indiv,
        nr_time = nr_time,
        i = 1:iter
      ) %>% 
      # Generate sets of person level parameters and data
      mutate(out = pmap(., simulate_twolevel_data)) %>% 
      unnest(out) %>% 
      nest(pop_pars = -c(i, nr_indiv, nr_time, indiv_pars, data))

    sims <- sims %>% 
      mutate(
        lmer = future_map(
          data,
          ~lmer(y ~ ylag_c + xlag_c + (ylag_c + xlag_c | individual), data = .x) %>% 
            parameters(effects = "all", group_level = TRUE) %>% 
            tibble() %>% 
            suppressMessages() %>% 
            suppressWarnings()
        )
      )

    sims <- sims %>% 
      mutate(
        brms = map(
          data,
          ~update(
            fit_brms_empty, 
            newdata = .x,
            cores = MAX_CORES,
            recompile = FALSE,
            threads = 2,
            iter = 2000
          ) %>% 
            parameters(centrality = "mean", effects = "all", group_level = TRUE, test = NULL) %>% 
            tibble()
        )
      )

    saveRDS(sims, path)
  } else {sims <- readRDS(path)}
  sims
}

# Run for specified sample sizes
sims <- tibble(people = 30, timepoints = c(6, 12, 18, 24, 30, 60)) %>% 
  mutate(
    out = map2(people, timepoints, ~simulate_estimate(.x, .y, iter = 100))
    ) %>% 
  unnest(out)
```

```{r}
#| label: fig-sim-res
#| fig-cap: Simulation study results. Each panel displays the mean posterior mean calculated over 100 iterations of model fitting. Vertical bars indicate standard errors over those 100 simulation runs. Each panel is labelled with a parameter name corresponding to the above equations. 
#| fig-width: 7

tmp <- sims %>% 
  mutate(
    result = pmap(
      list(lmer, brms, data),
      ~bind_rows(
        # Wrangle lmer results
        ..1 %>% 
          filter(Effects == "fixed") %>% 
          mutate(
            Parameter = fct_recode(
              Parameter, 
              `alpha[y]` = "(Intercept)",
              `beta[1]` = "ylag_c",
              `beta[2]` = "xlag_c"
            ),
            Centering = "Observed (lmer)"
          ),
        # Wrangle brms results
        ..2 %>% 
          filter(str_starts(Parameter, "b_|sd_|cor_|sigma")) %>% 
          mutate(
            Parameter = fct_recode(
              Parameter, 
              `alpha[x]` = "b_alphax_Intercept",
              `alpha[y]` = "b_alphay_Intercept",
              `beta[1]` = "b_betayy_Intercept",
              `beta[2]` = "b_betayx_Intercept",
              `tau[alpha[x]]` = "sd_individual__alphax_Intercept",
              `tau[alpha[y]]` = "sd_individual__alphay_Intercept",
              `tau[beta[1]]` = "sd_individual__betayy_Intercept",
              `tau[beta[2]]` = "sd_individual__betayx_Intercept",
              `rho[alpha[x]~alpha[y]]` = "cor_individual__alphax_Intercept__alphay_Intercept",
              `sigma[y]` = "sigma"
            )
          ) %>% 
          filter(!str_starts(Parameter, "cor_i")) %>% 
          mutate(Centering = "Latent (brms)") %>% 
          rename(Coefficient = Mean),
      ) %>% 
        bind_rows(
          # Observed mean of X
          ..3 %>% 
            summarise(Parameter = "alpha[x]", Coefficient = mean(x_b), Centering = "Observed (lmer)")
        )
    )
  ) %>% 
  select(1:4, result) %>% 
  unnest(result)
tmp %>%   
  ggplot(aes(nr_time, Coefficient, col = Centering, fill = Centering)) +
  scale_x_continuous(
    "Number of timepoints per person",
    breaks = extended_breaks()
  ) +
  scale_y_continuous(
    "Posterior mean",
    breaks = extended_breaks(),
    expand = expansion(.5)
  ) +
  geom_hline(
    data = pop %>% 
      select(
        `alpha[x]` = alpha_x, 
        `alpha[y]` = alpha_y, 
        `beta[1]` = beta_yy1, 
        `beta[2]` = beta_yx, 
        `tau[alpha[x]]` = tau_alpha_x, 
        `tau[alpha[y]]` = tau_alpha_y,
        `tau[beta[1]]` = tau_beta_yy1, 
        `tau[beta[2]]` = tau_beta_yx,
        `rho[alpha[x]~alpha[y]]` = cor_alpha, 
        `sigma[y]` = sigma_y
      ) %>% 
      pivot_longer(everything(), names_to = "Parameter", values_to = "Coefficient"),
    aes(yintercept = Coefficient),
    linewidth = .15
  ) +
  stat_summary(
    geom = "line", position = position_dodge(width = 2),
    fun = "mean", linewidth = .33
  ) +
  stat_summary(
    geom = "pointrange", position = position_dodge(width = 2),
    fun.data = "mean_se", fatten = 1.5, linewidth = .25
  ) +
  facet_wrap(
    "Parameter",
    nrow = 2,
    scales = "free_y",
    labeller = label_parsed
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = rel(1.1))
    )
```

>>>>

---
title: Estimating latent means model with regression models in R
author:
  - name: Matti Vuorre
    url: https://vuorre.com
    orcid: 0000-0001-5052-066X
    email: mjvuorre@uvt.nl
    affiliation:
      - ref: 1
  - name: Ethan McCormick
    affiliation:
      - ref: 3
  - name: Mauricio Garnier-Villarreal
    affiliation:
      - ref: 4
  - name: Joran Jongerling
    orcid: 0000-0001-5697-1381
    affiliation:
      - ref: 1
affiliations:
  - id: 1
    name: Tilburg University
  - id: 3
    name: todo
  - id: 4
    name: todo
authornote: |
  Work in progress.
abstract: |
  Researchers studying links between observed variables in longitudinal and other repeated measures data sets have long recognized the importance of person-mean centering in order to isolate within- and between-person associations. However, a critical issue in centering---treating the person means as unknown parameters rather than observed values---has received less attention among applied practitioners. One reason for this has been the general unavailability and perceived complexity of latent person-mean centering in contrast to observed mean centering. Here, we introduce latent person-mean centering and accessible syntax for implementing it in a wide variety of flexible bayesian models with the R package brms. We demonstrate the syntax, and benefits of latent centering, through an example analysis of social media effects on adolescent well-being. In appendices, we further contrast the performance of latent mean centering to observed mean centering via simulations, and show how it extends to more than two levels of analysis. Latent mean centering provides unbiased estimates of within- and between-person associations, and we hope that the availability and flexibility of our proposed approach motivates researchers to apply this method widely in longitudinal and other multilevel modelling scenarios.
keywords: "longitudinal data, repeated measures, centering, R, brms"
bibliography: references.bib

format: 
  html: default
  pdf: default
  docx: default
---

@@

```{r}
#| label: tab-data

d %>% 
  slice(1:3, .by = person) %>% 
  head() %>% 
  apa_table(
    caption = "Six rows of example data", 
    note = "sm = passive social media use; smg = grand-mean centered sm; smb = between-person sm; smw = within-person sm; happyw1 = within-person lagged happiness rating.",
    col_spanners = list(`Original data` = c(1, 5), `Centered values` = c(6, 9)),
    digits = c(0, 0, 0, 0, 0, 0, 1, 1, 1), 
    font_size = "small", 
    span_text_columns = FALSE
  )
```

@@

@@

Conceptualization: MV  
Methodology: MV  
Software: MV  
Validation: MV  
Formal Analysis: MV  
Data Curation: MV  
Writing---Original Draft: MV  
Writing---Review & Editing: MV  
Visualization: MV  
Project Administration: MV

@@
